---
title: "Salt Lake County Neighborhoods: The Right Investment for Juvenile Justice Success"
subtitle: "Transforming communities through strategic investment in youth and families rather than incarceration"
author: "Based on Utah's 2025 Statewide Youth Geospatial Study"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    css: styles.css
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(plotly)
library(knitr)
library(kableExtra)
library(DT)
library(leaflet)
library(htmltools)

# Set plot theme
theme_set(theme_minimal() + 
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.title = element_text(size = 11),
        legend.position = "bottom"))
```

```{css}
#| echo: false
body {
  font-family: 'Segoe UI', Tahoma, Geneva, Verdana, sans-serif;
  line-height: 1.6;
  color: #333;
}

.hero-section {
  background: linear-gradient(135deg, #2c5aa0 0%, #1e3a5f 100%);
  color: white;
  padding: 3rem 2rem;
  margin-bottom: 2rem;
  border-radius: 8px;
}

.hero-section h1 {
  font-size: 2.5rem;
  margin-bottom: 1rem;
  font-weight: 700;
}

.hero-section .subtitle {
  font-size: 1.2rem;
  opacity: 0.9;
  margin-bottom: 2rem;
}

.cost-highlight {
  background: #f8f9fa;
  border-left: 4px solid #dc3545;
  padding: 1.5rem;
  margin: 2rem 0;
  border-radius: 4px;
}

.investment-highlight {
  background: linear-gradient(135deg, #28a745 0%, #20c997 100%);
  color: white;
  padding: 1.5rem;
  margin: 2rem 0;
  border-radius: 8px;
}

.neighborhood-card {
  background: white;
  border: 1px solid #dee2e6;
  border-radius: 8px;
  padding: 1.5rem;
  margin: 1rem 0;
  box-shadow: 0 2px 4px rgba(0,0,0,0.1);
}

.stats-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(200px, 1fr));
  gap: 1rem;
  margin: 2rem 0;
}

.stat-card {
  background: #f8f9fa;
  padding: 1.5rem;
  text-align: center;
  border-radius: 8px;
  border: 2px solid #e9ecef;
}

.stat-number {
  font-size: 2.5rem;
  font-weight: bold;
  color: #2c5aa0;
  display: block;
}

.stat-label {
  font-size: 0.9rem;
  color: #6c757d;
  margin-top: 0.5rem;
}

.alternative-programs {
  background: #e3f2fd;
  padding: 2rem;
  border-radius: 8px;
  margin: 2rem 0;
}

.program-cost {
  display: flex;
  justify-content: space-between;
  align-items: center;
  padding: 1rem;
  background: white;
  margin: 0.5rem 0;
  border-radius: 4px;
  border-left: 4px solid #28a745;
}

.protective-factors {
  background: #fff3cd;
  border: 1px solid #ffeaa7;
  padding: 1.5rem;
  border-radius: 8px;
  margin: 2rem 0;
}

.tier-1 { border-left: 4px solid #dc3545; }
.tier-2 { border-left: 4px solid #fd7e14; }
.tier-3 { border-left: 4px solid #ffc107; }
.tier-4 { border-left: 4px solid #17a2b8; }

.plotly-container {
  margin: 2rem 0;
}
```

```{r data-setup}
#| include: false

# Sample data for demonstration - replace with actual data from your analysis
utah_stats <- data.frame(
  metric = c("Court Referrals per 1,000 Youth", "Utah Incarceration Rate per 100,000", 
             "Diversion Success Rate (%)", "Reduction in Referrals (%)"),
  value = c(3.4, 37, 94, 35),
  comparison = c("Serious offenses FY24", "vs. 74 nationally", 
                "Complete programs successfully", "Between FY17-FY24")
)

# Program cost data
program_costs <- data.frame(
  program = c("Multisystemic Therapy (MST)", "Functional Family Therapy (FFT)", 
              "Community Mentorship", "Mobile Crisis Teams", 
              "After-School Programs", "Job Training"),
  cost_per_unit = c(8000, 5500, 2500, 150000, 1200, 4000),
  unit = c("family", "family", "youth", "team", "youth", "participant"),
  potential_served = c(1875, 2727, 6000, 100, 12500, 3750),
  description = c("Targets family functioning, school, and community support",
                  "In-home therapeutic services for at-risk youth",
                  "One-on-one support and guidance", 
                  "24/7 mental health crisis intervention",
                  "Safe spaces with academic and recreational activities",
                  "Workforce development for older youth")
)

# Priority neighborhoods data
priority_neighborhoods <- data.frame(
  tier = rep(c("Tier 1", "Tier 2"), c(5, 5)),
  neighborhood = c("Glendale/Poplar Grove", "West Valley City Central", 
                   "Rose Park/Jordan River", "Kearns/Taylorsville West",
                   "South Salt Lake East", "Additional West Valley", 
                   "Midvale/Murray Border", "Additional Kearns", 
                   "West Jordan Zones", "Magna/West Valley"),
  risk_factors = c("High referral rates, spatial clustering", 
                   "Population density, elevated episodes",
                   "County-relative elevated risk", "Multi-factor concentration",
                   "Emerging hot spot", "Elevated indicators", 
                   "Border area challenges", "Additional risk tracts",
                   "Targeted zones", "Corridor challenges"),
  composite_score = c(95, 92, 88, 85, 82, 78, 75, 72, 68, 65)
)

# Expected outcomes data
expected_outcomes <- data.frame(
  outcome = c("Reduction in Justice Involvement", "ROI on Prevention", 
              "Family Satisfaction", "School Engagement Improvement"),
  percentage = c("50-70%", "$7:1", "85%+", "25-40%"),
  description = c("Reduction in youth justice involvement", 
                  "Return on investment in prevention programs",
                  "Family satisfaction with community services",
                  "Improvement in school engagement")
)
```

::: {.hero-section}
# Salt Lake County Neighborhoods: The Right Investment for Juvenile Justice Success

::: {.subtitle}
Transforming communities through strategic investment in youth and families rather than incarceration
:::

**Based on Utah's 2025 Statewide Youth Geospatial Study findings**
:::

## Executive Summary

Salt Lake County taxpayers spend approximately **$15-25 million annually** to process and supervise youth in the juvenile justice system. The burden of juvenile justice involvement is disproportionately felt by specific neighborhoods within the county. These same communities that experience the highest rates of juvenile justice involvement also face the greatest challenges serving their residents.

Rather than spend millions of dollars to process and detain youth from these communities, these funds could instead be invested in education, mentorship, mental health services, family support, and youth development programs that address root causes and build community strength.

::: {.cost-highlight}
**Current Reality**: Youth incarceration costs an average of $588 per day per child, while community-based programs can cost as little as $75 per day.

**The Opportunity**: Utah's early intervention services have achieved a 98% success rate, with youth avoiding new custody and probation supervision at 180 days follow-up.
:::

## Current System Performance

```{r utah-stats-chart}
#| fig-cap: "Key Utah Juvenile Justice Statistics"

p <- utah_stats %>%
  ggplot(aes(x = reorder(metric, value), y = value)) +
  geom_col(fill = "#2c5aa0", alpha = 0.8) +
  geom_text(aes(label = paste0(value, ifelse(grepl("%", metric), "%", ""))), 
            hjust = -0.1, color = "#333", size = 4, fontface = "bold") +
  coord_flip() +
  labs(title = "Utah Juvenile Justice System Performance",
       subtitle = "Current statistics showing system effectiveness",
       x = "", y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12))

ggplotly(p, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

The cost of justice system involvement includes not just the direct financial burden, but also the social and emotional costs to youth and their families:

- **Lost educational opportunities**: Youth incarceration significantly increases a child's likelihood of developing an emotional or behavioral disorder
- **Increased risk of recidivism**: Living in a juvenile justice institution increases the probability of a teen becoming arrested as an adult by 22-26%
- **Economic impact**: We spend 1250% more to incarcerate a child than to educate them
- **Family destabilization**: Court costs, fines, fees, and restitution hurt youth and families

## Priority Neighborhoods for Investment

```{r priority-neighborhoods-chart}
#| fig-cap: "Priority Neighborhoods by Composite Risk Score"

p2 <- priority_neighborhoods %>%
  ggplot(aes(x = reorder(neighborhood, composite_score), y = composite_score, 
             fill = tier)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Tier 1" = "#dc3545", "Tier 2" = "#fd7e14")) +
  coord_flip() +
  labs(title = "Salt Lake County Priority Neighborhoods",
       subtitle = "Ranked by composite risk score for targeted investment",
       x = "Neighborhood", y = "Composite Risk Score", fill = "Priority Tier") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.position = "top")

ggplotly(p2, tooltip = c("y", "fill")) %>%
  layout(showlegend = TRUE)
```

Based on Utah's comprehensive geospatial analysis combining court referral rates, hot spot analysis, county comparisons, and population impact, the neighborhoods above represent the highest opportunity areas for strategic investment.

```{r priority-table}
#| tbl-cap: "Detailed Priority Neighborhood Analysis"

priority_neighborhoods %>%
  select(Tier = tier, Neighborhood = neighborhood, 
         `Risk Factors` = risk_factors, `Composite Score` = composite_score) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(which(priority_neighborhoods$tier == "Tier 1"), 
           background = "#ffebee") %>%
  row_spec(which(priority_neighborhoods$tier == "Tier 2"), 
           background = "#fff3e0")
```

## Community-Specific Risk and Protective Factors

```{r risk-factors-table}
#| tbl-cap: "Key Indicators Affecting Juvenile Justice Success"

risk_factors <- data.frame(
  Indicator = c("Economic Stability", "Educational Access", "Family Support", 
                "Mental Health Access", "Transportation", "Housing Stability"),
  `High-Risk Areas` = c("Areas with >25% poverty rates", "Schools with >20% chronic absenteeism",
                        "Single-parent households >40%", "Limited provider availability",
                        "Limited public transit access", ">15% rental turnover annually"),
  `Protective Factors` = c("Strong local employment hubs", "High-performing neighborhood schools",
                           "Strong community organizations", "Integrated community health centers",
                           "Walkable neighborhoods with resources", "Stable homeownership rates")
)

risk_factors %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## The Investment Opportunity

::: {.investment-highlight}
### What Salt Lake County Could Fund Instead

**For the annual cost of juvenile justice processing in priority neighborhoods, Salt Lake County could invest in:**
:::

```{r program-costs-chart}
#| fig-cap: "Alternative Program Investment Options"

p3 <- program_costs %>%
  mutate(program_label = paste0(program, " ($", format(cost_per_unit, big.mark = ","), 
                               " per ", unit, ")")) %>%
  ggplot(aes(x = reorder(program_label, potential_served), y = potential_served)) +
  geom_col(fill = "#28a745", alpha = 0.8) +
  geom_text(aes(label = format(potential_served, big.mark = ",")), 
            hjust = -0.1, color = "#333", size = 3.5, fontface = "bold") +
  coord_flip() +
  labs(title = "Community Program Investment Capacity",
       subtitle = "Number of individuals/families that could be served annually",
       x = "Program Type", y = "Potential Individuals Served") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text.y = element_text(size = 10))

ggplotly(p3, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

```{r program-details-table}
#| tbl-cap: "Evidence-Based Community Program Details"

program_costs %>%
  select(Program = program, `Cost per Unit` = cost_per_unit, 
         Unit = unit, `Potential Served` = potential_served, 
         Description = description) %>%
  mutate(`Cost per Unit` = paste0("$", format(`Cost per Unit`, big.mark = ","))) %>%
  mutate(`Potential Served` = format(`Potential Served`, big.mark = ",")) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

::: {.protective-factors}
### Strength-Based Approach: Building on What Works

**Key Questions for Community Investment:**
- What strengths does each community already have?
- Where are people already gathering, participating, or supporting one another?
- What protective factors can be enhanced?

#### Identified Protective Factors in Salt Lake County:
- **Strong family connections** in immigrant and refugee communities
- **Faith-based organization networks** providing community support
- **Neighborhood schools** serving as community anchors
- **Local businesses** providing informal mentorship and employment
- **Cultural organizations** maintaining community identity and pride
- **Community gardens and recreation centers** fostering social connections
:::

## Evidence-Based Strategies That Work

Research demonstrates that the biggest reductions in youth delinquency happen in programs that are community-based, target multiple needs, and improve caregiver skills for communication and monitoring.

### Proven Approaches Include:

```{r evidence-chart}
#| fig-cap: "Success Rates of Evidence-Based Interventions"

evidence_data <- data.frame(
  intervention = c("Early Intervention Services", "Diversion Programs", 
                   "Multisystemic Therapy", "Community-Based Treatment"),
  success_rate = c(98, 94, 85, 80),
  description = c("Utah's early intervention services", 
                  "Youth diverted from formal court processing",
                  "Family-focused therapeutic intervention",
                  "Community-based vs. residential treatment")
)

p4 <- evidence_data %>%
  ggplot(aes(x = reorder(intervention, success_rate), y = success_rate)) +
  geom_col(fill = "#20c997", alpha = 0.8) +
  geom_text(aes(label = paste0(success_rate, "%")), 
            hjust = -0.1, color = "#333", size = 4, fontface = "bold") +
  coord_flip() +
  labs(title = "Evidence-Based Intervention Success Rates",
       subtitle = "Percentage of youth with positive outcomes",
       x = "Intervention Type", y = "Success Rate (%)") +
  ylim(0, 105) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12))

ggplotly(p4, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## Expected Outcomes

Based on evidence from similar investments nationwide:

```{r outcomes-chart}
#| fig-cap: "Projected Outcomes from Community Investment"

# Create a radar chart effect with ggplot
outcomes_numeric <- data.frame(
  outcome = c("Justice Involvement Reduction", "ROI Multiple", 
              "Family Satisfaction", "School Engagement"),
  min_value = c(50, 5, 80, 25),
  max_value = c(70, 9, 90, 40),
  target = c(60, 7, 85, 32)
)

p5 <- outcomes_numeric %>%
  ggplot() +
  geom_segment(aes(x = reorder(outcome, target), y = min_value, 
                   xend = reorder(outcome, target), yend = max_value), 
               color = "#6c757d", size = 3, alpha = 0.6) +
  geom_point(aes(x = reorder(outcome, target), y = target), 
             color = "#2c5aa0", size = 6) +
  geom_text(aes(x = reorder(outcome, target), y = target + 5, 
                label = paste0(target, ifelse(outcome == "ROI Multiple", ":1", "%"))), 
            color = "#333", size = 4, fontface = "bold") +
  coord_flip() +
  labs(title = "Expected Outcomes from Community Investment",
       subtitle = "Projected improvements based on national evidence",
       x = "Outcome Measure", y = "Percentage/Multiple") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12))

ggplotly(p5, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## Regional Implementation Strategy

### Wasatch Front Coordination
Given that priority neighborhoods span multiple jurisdictions, a coordinated regional approach maximizes impact:

- **Salt Lake County**: Lead coordination and resource allocation
- **Municipal Partners**: West Valley City, South Salt Lake, Kearns
- **School Districts**: Granite, Jordan, Murray school districts  
- **Community Organizations**: Neighborhood houses, cultural centers, faith communities
- **Healthcare Partners**: Community health centers, mental health providers

### Phased Implementation

```{r implementation-timeline}
#| fig-cap: "Implementation Timeline and Phases"

timeline_data <- data.frame(
  phase = rep(c("Phase 1", "Phase 2", "Phase 3"), each = 3),
  activity = c("Deploy mobile crisis teams", "Establish resource centers", "Launch family therapy",
               "Scale mentorship programs", "Implement job training", "Strengthen partnerships", 
               "Measure outcomes", "Build funding partnerships", "Expand successful models"),
  start_month = c(1, 2, 3, 7, 9, 12, 19, 24, 30),
  duration = c(6, 6, 6, 12, 12, 12, 18, 12, 12)
)

timeline_data$end_month <- timeline_data$start_month + timeline_data$duration

p6 <- timeline_data %>%
  ggplot() +
  geom_segment(aes(x = start_month, y = reorder(activity, -start_month), 
                   xend = end_month, yend = reorder(activity, -start_month),
                   color = phase), size = 4, alpha = 0.8) +
  scale_color_manual(values = c("Phase 1" = "#dc3545", "Phase 2" = "#ffc107", "Phase 3" = "#28a745")) +
  labs(title = "Implementation Timeline",
       subtitle = "Phased approach to community investment",
       x = "Months", y = "Activities", color = "Implementation Phase") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.position = "top")

ggplotly(p6, tooltip = c("colour", "x")) %>%
  layout(showlegend = TRUE)
```

## Call to Action

Salt Lake County stands at a critical juncture. Utah's 2025 Statewide Geospatial Study provides communities with data to support youth and families while lowering court referral rates. The data clearly shows where investment is needed most.

By redirecting resources from reactive processing to proactive community investment, Salt Lake County can:

- **Strengthen families** and keep them together
- **Build safer communities** through prevention rather than punishment  
- **Generate significant cost savings** while improving outcomes
- **Address root causes** of youth justice involvement
- **Create lasting positive change** for generations to come

### Next Steps

```{r next-steps-checklist}
next_steps <- data.frame(
  step = 1:5,
  action = c("Community Engagement", "Resource Reallocation", "Provider Network Development",
             "Data Tracking", "Policy Development"),
  description = c("Partner with priority neighborhoods to understand needs and strengths",
                  "Begin shifting funding from reactive to preventive approaches", 
                  "Establish partnerships with evidence-based program providers",
                  "Implement systems to measure progress and outcomes",
                  "Create supportive policies for community-based interventions"),
  timeline = c("Immediate", "0-6 months", "3-9 months", "6-12 months", "9-18 months")
)

next_steps %>%
  select(Step = step, Action = action, Description = description, Timeline = timeline) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

---

*For the full Utah study and additional data, visit: [https://justice.utah.gov/research-analysis/2025-statewide-youth-geospatial-study/](https://justice.utah.gov/research-analysis/2025-statewide-youth-geospatial-study/)*

*This analysis is based on data from Utah's Commission on Criminal and Juvenile Justice, the Juvenile Justice Oversight Committee, and evidence-based research from national juvenile justice organizations.*

```{r session-info}
#| include: false
#| eval: false
# Session information for reproducibility
sessionInfo()
```
